---
title: "AIR PASSENGERS PREDICTION MODEL"
author: "Enock Bereka"
date: "2025-01-25"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Air Passengers Prediction
# Install and load necessary packages
```{r warning=FALSE, message=FALSE}
library(forecast)
library(tseries)
library(ggplot2)
```

## Load the AirPassengers dataset
```{r warning=FALSE, message=FALSE}
data("AirPassengers")
```

## Data Exploration and Preprocessing
```{r warning=FALSE, message=FALSE}
str(AirPassengers)
attributes(AirPassengers)
```

## Plot the time series data
```{r warning=FALSE, message=FALSE}
plot(AirPassengers, 
     main = "Monthly Airline Passengers (1949-1960)",
     xlab = "Year", ylab = "Passengers")
abline(reg = lm(AirPassengers ~ time(AirPassengers)))
```

```{r warning=FALSE, message=FALSE}
## General trend
plot(aggregate(AirPassengers, FUN = mean))
```
# There is a general increase in the number of passengers from 1949-1960

## Decomposition of the Time series
```{r warning=FALSE, message=FALSE}
decomposition <- decompose(AirPassengers)
plot(decomposition)
```
## From the above figure, in the trend part we can see that there is upward trend and from the seasonal part we can see that there is reoccurring seasonal patterns (which indicates seasonality).

## Check for Stationarity
# Plot the ACF plot
```{r warning=FALSE, message=FALSE}
acf(AirPassengers)
```

# Plot the PACF plot
```{r warning=FALSE, message=FALSE}
pacf(AirPassengers)
```
## From the ACF and PACF plot we can see that the vertical lines crossing the blue horizontal line which
indicates the high correlation between the present value sand its lagged versions. Therefore, the 
time series is not stationary 

## Perform ADF test
```{r warning=FALSE, message=FALSE}
adf.test(AirPassengers, k = 15)
```
## P value is more than 0.05 indicating that the data is not stationary

## Perform KPSS test
```{r warning=FALSE, message=FALSE}
kpss.test(AirPassengers)
```
## P value is less than 0.05 indicating that the data is not stationary
## We will not do differencing and log transformation since it will be automatically be handled during model building using auto.arima.

## Model building
```{r warning=FALSE, message=FALSE}
arima_model <- auto.arima(AirPassengers)
summary(arima_model)
```

## Perform residual diagnostic analysis
```{r warning=FALSE, message=FALSE}
acf(ts(arima_model$residuals))
```
```{r warning=FALSE, message=FALSE}
pacf(ts(arima_model$residuals))
```
# From the above ACF and PACF plots we can infer that all the vertical lines at each lags falls inside the horizontal boundary, which indicates that the time series is stationary after applying seasonal differencing using the auto.arima model. 

## Plot a histogram of residuals
```{r warning=FALSE, message=FALSE}
residuals <- residuals(arima_model)
hist(residuals, main="Histogram of Residuals", 
     xlab="Residuals", col="skyblue", border="black", 
     breaks=20, probability=TRUE)
# Add a normal density curve
lines(density(residuals), col="red", lwd=2)
```
# The histogram indicates that residuals are normally distributed

## Forecasting
```{r warning=FALSE, message=FALSE}
forecast_future <- forecast(arima_model, level = c(95), h = 96)
forecast_future
```
## The model predicts a general increase in the number of passengers as the number years increase

## Visualize your Predictions
```{r warning=FALSE, message=FALSE}
plot(forecast_future)
```
```{r warning=FALSE, message=FALSE}
forecast_df <- data.frame(
  Date = time(forecast_future$mean),
  Forecast = as.numeric(forecast_future$mean),
  Lower = as.numeric(forecast_future$lower[, 1]),
  Upper = as.numeric(forecast_future$upper[, 1]))

# Plot the forecast using ggplot2
ggplot(forecast_df, aes(x = Date, y = Forecast)) +
  geom_line(color = "red") +
  theme_classic()+
  geom_ribbon(aes(ymin = Lower, ymax = Upper), alpha = 0.2) +
  labs(title = "Air Passengers Forecast", x = "Year", y = "TB Incidence")
```

## Goodness of fit of the model
#Ljung-Box test
```{r warning=FALSE, message=FALSE}
Box.test(forecast_future$residuals, type = "Ljung-Box")
```
## From the results of Ljung-Box test, we can infer that the p-value is greater than 0.05. Therefore,
the given model is a good-fit.


